# Icosahedron Earth {#sec-dodecahedron}

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

# Package imports
library(ggplot2)
library(kableExtra)

# Set seed for reproducibility
set.seed(12345)
```

Of course, the Earth isn't a tetrahedron, but an [oblate spheroid](https://en.wikipedia.org/wiki/Figure_of_the_Earth), and there are an infinite number of points we could measure on its surface, not just four distinct faces. To get closer to this shape, we're going to change our model from a tetrahedron to another regular Platonic solid: am icosahedron, with 20 faces.

::: {.column-margin}

![The Earth, projected onto an icosahedron. Image by [John D. Cook](https://www.johndcook.com/blog/2018/08/22/globe-regular-solids/)](assets/images/world_icosa.png){#fig-world-icosa}

:::

::: {.callout-note}
Changing our model of the Earth to **Icosahedron Earth**, with 20 faces, is still artificial. But it will allow us to see the shape of the probability distribution as we change the number of samples we take, more clearly.
:::

## Calculating probabilities

We define that **Icosahedron Earth** has 20 faces, with each face being either "water" (`W`) or "land (`L`). This implies that there are 21 possibilities, from "all water," through "one land, 19 water," all the way to "all land." It would take us quite a while to calculate out and show the possibilities explicitly. Happily, being in the age of modern computational statistics we can use [the `R` language](https://www.r-project.org/)[^1] to do the calculations for us and visualise the result.

[^1]: This entire online book is written in the `R` language. It's rather powerful and flexible.

### Testing Tetrahedron Earth

The code below calculates these probabilities for our observations in @sec-wet-earth, on Tetrahedron Earth:

> `L W L L W W W L W W`

and reproduces the values from @fig-ten-obs-prob in @tbl-tetra-post. 

```{r}
#| label: tbl-tetra-post
#| tbl-cap: For Tetrahedron Earth, each possibile proportion of surface water, the corresponding ways the observed outcome could be generated, and the _posterior probability_ that the Tetrahedron Earth we sampled from has that proportion of surface water.
#| html-table-processing: none

# Return the posterior distribution of p, the proportion of points (faces) 
# on the planet's surface that are water, given a sample in the form
# c("W", "L", "W") and a number of faces.
# By default the number of faces is four, representing Tetrahedron Earth.
# Adapted from Statistical Rethinking, by Richard McElreath
# (https://www.youtube.com/watch?v=R1vcdhPBlXA)
compute_posterior <- function(sample, faces=4, round_to=3) {
  W <- sum(sample == "W")  # Waters observed
  L <- sum(sample == "L")  # Lands observed
  possibilities = faces + 1  # Number of possible probabilities
  proportions = seq(0, 1, length.out=possibilities)  # The possibile proportions of water
  
  # Calculate the number of ways the sample data could be generated
  ways <- sapply(proportions, function(q) (q * faces) ^ W * ( (1-q) * faces ) ^ L )
  
  # Calculate the posterior distribution
  posterior <- ways/sum(ways)
  
  # Return a data.frame of the possibilities and ways
  data.frame( proportion=factor(proportions), ways,
              probability=round(posterior, round_to))
}

# The observations we have already made
observations <- c("L", "W", "L", "L", "W", "W", "W", "L", "W", "W")

# Calculate the resulting distribution
distn_tetra <- compute_posterior(observations, 4)

# Tabulate the result
distn_tetra |>
  kbl(format="html", align="c") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

As we are using modern computational statistics in `R`, we can also visualise the distribution using the code below, in @fig-tetra-post.

```{r}
#| label: fig-tetra-post
#| fig-cap: A histogram of the probabilities that each possible form of Tetrahedron Earth could have produced the observed sequence of observations.
#| fig-width: 6
#| fig-column: margin

# Show the Tetrahedron Earth distribution as a histogram
ggplot(distn_tetra, aes(x=proportion, y=probability)) +
  geom_col(fill="darkolivegreen4") +
  theme_light() +
  theme(text = element_text(size=20))
```

We can see visually and intuitively in @fig-tetra-post that the _posterior probability distribution_ skews to the right, and there is not much to choose between the 50% and 75% water planets.

::: {.callout-important}
Our analysis here, as is the case for traditional _frequentist_ analysis, does not give us a single point value as output to tell us "this form of **Tetrahedron Earth** is the one that generate the observed sample data. It gives us a _probability distribution_ that we then must interpret.

You are probably used to interpreting the outputs of _t_-tests and similar statistical analyses in terms of whether a probability is greater or less than some threshold value. It is important to recognise that this probability comes from a _distribution_ of possibilities as well. The difference here is that we are making that distribution explicit, and not applying a threshold.
:::

### Probabilities for Icosahedron Earth

We wrote the code above to calculate _posterior probabilities_ in `R`, and so we can reuse the code to investigate circumstances different to Tetrahedron Earth. Specifically, we are interested to know what happens if we have a "more realistic" (but still highly artificial) model: **Icosahedron Earth**, with 20 faces that are either "water" or "land". We can calculate this using the code below.

```{r}
#| label: fig-icosa-post
#| fig-cap: A histogram of the probabilities that each possible form of Icosahedron Earth (with 20 faces) could have produced the observed sequence of observations.
#| fig-width: 6

# Calculate posterior probabilities for Icosahedron Earth
distn_icosa <- compute_posterior(observations, faces=20)

# Show the Tetrahedron Earth distribution as a histogram
ggplot(distn_icosa, aes(x=proportion, y=probability)) +
  geom_col(fill="darkolivegreen4") +
  theme_light() +
  theme(text = element_text(size=10))
```

Visually we can now see that the distribution is much _smoother_ and starts to resemble the _parametric probability distributions_ you might already be familiar with.


::: {.callout-tip}
## Interpreting the distribution

The distribution in @fig-icosa-post is much more informative than that in @fig-tetra-post. The overall shape of the probability distribution is clearer, and we can see it looks like a skewed bell shape, with a _modal_[^2] value of 0.6, implying that the most likely surface water distribution is 60% of the surface of **Icosahedron Earth**.

We can also see that the bulk of the estimates lie between 0.45 and 0.75, suggesting that the analysis is most _confident_ that the true distribution lies somewhere between these values[^3].
:::

[^2]: Most common
[^3]: This leads to the concept of _confidence intervals_ and _credibility intervals_, which we don't have time to cover.

::: {.callout-important}
**We haven't increased the number of samples we have taken or values we have observed**

We have only made our _model Earth_ more representative of the "real" situation of an oblate spheroid Earth.
:::

::: {.callout-warning}
Although the estimate of the most likely value for $p$ has become more precise, the _y_-axis tells us that the **probability of any specific value of $p$ being the "true" value has fallen**, and the relative probability of nearby possible values for $p$ is very similar.
:::

### What if we used a larger number of faces?

We might think that if we continue to increase the level of detail in our model, we will keep increasing the precision of our estimate. We can see what the distribution would look like for a model Earth with 100 faces, using the code below.

```{r}
#| label: fig-100-post
#| fig-cap: A histogram of the posterior probabilities, using an Earth with 100 faces.
#| fig-width: 6

# Calculate posterior probabilities for Icosahedron Earth
distn_hundo <- compute_posterior(observations, faces=100, round_to=5)

# Show the Tetrahedron Earth distribution as a histogram
ggplot(distn_hundo, aes(x=proportion, y=probability)) +
  geom_col(fill="darkolivegreen4") +
  theme_light() +
  theme(text = element_text(size=10)) +
  scale_x_discrete(breaks=seq(0, 1, 0.1))
```

We do see that the estimate of the most likely proportion $p$ is more _precise_, but the relative probability of nearby values is also very similar.

::: {.callout-important}
When changing the number of faces used to represent the Earth, there is a trade-off between the _precision_ of our estimate of the most likely proportion $p$, and our _confidence_ in that estimate being the most likely proportion.
:::
